%\documentstyle[11pt,psfig,fullpage,url]{report}
\documentclass[11pt]{report}
\usepackage{fullpage}
%\usepackage{psfig}
\usepackage{graphicx}
\usepackage{url,cleveref}

\setlength{\topmargin}{-0.5cm}

\newcommand{\fig}[1]{Figure~\ref{fig:#1}}
%\renewcommand{\sec}[1]{Section~\ref{sec:#1}} % This destroys \sec for secant.
\newcommand{\tab}[1]{Table~\ref{tab:#1}}
\newcommand{\ignore}[1]{}

\newcommand{\superlu}{{\tt SuperLU}}              % "SuperLU"
\newcommand{\superlud}{{\tt SuperLU\_DIST}}       % "SuperLU_DIST"
\newcommand{\superlumt}{{\tt SuperLU\_MT}}        % "SuperLU_DIST"
\newcommand{\metis}{{\tt MeTiS}}
\newcommand{\parmetis}{{\tt ParMeTiS}}

\def\SuperLU{\mbox{SuperLU}}
%\def\superlumt{\mbox{SuperLU\_MT}}
%\def\superlud{\mbox{SuperLU\_DIST}}
\def\BLAS{\mbox{BLAS}}
\def\LAPACK{\mbox{LAPACK}}

\def\NC{\mbox{\tt SLU\_NC}}
\def\NR{\mbox{\tt SLU\_NR}}
\def\NRloc{\mbox{\tt SLU\_NR\_loc}}
\def\NCP{\mbox{\tt SLU\_NCP}}
\def\SC{\mbox{\tt SLU\_SC}}
\def\SCP{\mbox{\tt SLU\_SCP}}
\def\DN{\mbox{\tt SLU\_DN}}
\def\GE{\mbox{\tt SLU\_GE}}
\def\TRLU{\mbox{\tt SLU\_TRLU}}
\def\TRU{\mbox{\tt SLU\_TRU}}
\def\S{\mbox{\tt SLU\_S}}
\def\D{\mbox{\tt SLU\_D}}
\def\C{\mbox{\tt SLU\_C}}
\def\Z{\mbox{\tt SLU\_Z}}

\def\ptt{\tt}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\title{\SuperLU\ Users' Guide%
\thanks{The work was supported by Director, Office of Science,
Office of Advanced Scientic Computing Research of the U.S. Department of
Energy under Contract No. DE-AC02-05CH11231.
Additional support was provided by the U.S. Dept. of Energy (DOE), Office of Science,
Advanced Scientific Computing Research
(and Basic Energy Sciences/Biological and Environmental Research/High
Energy Physics/Fusion Energy Sciences/Nuclear Physics),
through the Scientific Discovery through Advanced Computing (SciDAC) program.}
}

\author{
  \hskip.5in\mbox{\ }
        Xiaoye S. Li%
	\thanks{
        Lawrence Berkeley National Laboratory, MS 50F-1650, 1 Cyclotron Rd,
        Berkeley, CA 94720. (xsli@lbl.gov).
%  This work was supported
%  in part by the Director, Office of Advanced Scientific Computing
%       Research of the U.S. Department of Energy under Contract
%        No. D-AC02-05CH11231.
	}
\and
        James W. Demmel%
        \thanks{Computer Science Division, University of California,
        Berkeley, CA 94720. (demmel@cs.berkeley.edu).
        The research of Demmel and Li was supported in part by 
        NSF grant ASC--9313958,
        DOE grant DE--FG03--94ER25219,
        UT Subcontract No.\ ORA4466 from ARPA Contract No.\ DAAL03--91--C0047,
        DOE grant DE--FG03--94ER25206,
        and
        NSF Infrastructure grants CDA--8722788 and CDA--9401156.
        }
\and
        John R. Gilbert%
        \thanks{
 	Department of Computer Science, University of California,
	 Santa Barbara, CA 93106. (gilbert@cs.ucsb.edu).
%	Xerox Palo Alto Research Center,
%	3333 Coyote Hill Road,
%        Palo Alto, CA 94304 (gilbert@parc.xerox.com).
        The research of this author was supported in part by the
	Institute for Mathematics and Its Applications at the University
	of Minnesota and in part by DARPA Contract No. DABT63-95-C0087.
	Copyright $\copyright$ 1994-1997 by Xerox Corporation.
        All rights reserved.}
\and
    Laura Grigori%
    \thanks{
    INRIA Saclay-Ile de France,
    Laboratoire de Recherche en Informatique,
    Universite Paris-Sud 11. (laura.grigori@inria.fr)
    }
\and 
    Piyush Sao%
    \thanks{
      Computational Science and Engineering Division,
      Georgia Institute of Technology, Atlanta.
      (piyush3@gatech.edu).
    The research of this author was supported in part by the
    National Science Foundation CAREER award number 0953100 and
    DOE X-Stack 1.0 under DE-FC02-10ER26006/DE-SC0004915 (PI: Richard Vuduc).
    }
\and
    Meiyue Shao%
    \thanks{
      Computational Research Division,
      Lawrence Berkeley National Laboratory,
      MS 50F-1650, 1 Cyclotron Rd,
      Berkeley, CA 94720. (myshao@lbl.gov).
%    Department of Computing Science and HPC2N,
%    Ume{\aa} University,
%    SE-901 87, Ume{\aa}, Sweden. (myshao@cs.umu.se)
    }
\and
    Ichitaro Yamazaki%
    \thanks{
    Innovative Computing Laboratory,
    Department of Electrical Engineering and Computer Science,
    The University of Tennessee. (ic.yamazaki@gmail.com).
    The research of this author was supported in part by the Director,
    Office of Advanced Scientific Computing Research of the
    U.S. Department of Energy under Contract No. D-AC02-05CH11231.}
}


%\date{November 15, 1997}
%\date{December 16, 2002}
\date{September 1999 \\
% Last update: March 2003 % SuperLU_DIST v2.0 - distributed input interface
% Last update: September 2003 % fixed the MPI buffer bug in triangular solve
% Last update: October 2003 % SuperLU v3.0
% Last update: September 2007 % SuperLU_MT v2.0
% Last update: November 2007 % SuperLU_DIST v2.1
% Last update: June 2009 % SuperLU v4.0
% Last update: October 2009
% Last update: August 2011   % SuperLU_DIST v3.0
\vspace{4mm}
% Last update: October 2014   % SuperLU_DIST v4.0, included GPU, OpenMP for factorization
%Last update: March 2016   % Frank MCKENNA: Manual has 1 more integer arg for
                           % LUstructInit(m,n,&LUstruct)
Last update: June 2018}   % Added parallel AWPM row pivoting option

\begin{document}
\maketitle
\tableofcontents

\renewcommand{\thefootnote}{\arabic{footnote}}

\chapter{Introduction}
\label{chap:intro}
\input{intro}

\chapter{Sequential SuperLU (Version 4.2)}
\label{chap:superlu}
\input{ug_part1}

%\chapter{Multithreaded SuperLU (Version 1.1)}
\chapter{Multithreaded SuperLU (Version 2.0)}
\label{chap:superlu_mt}
\input{ug_part2}

\chapter{Distributed-memory SuperLU on manycore nodes (Version 4.0)}
\label{chap:superlu_dist}
\input{ug_part3}

%\newpage
\bibliographystyle{plain}
\bibliography{sparse,others}

% -----------------------------------------------------------
%\input{append}

\end{document}
